---
title: "Validation for initial submission"
author: "XSun"
date: "2025-07-31"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

## Enrichment in other gene sets

https://pubmed.ncbi.nlm.nih.gov/31981491/ This paper published several data sets we can use. The gene sets are gtex tissue specific genes, prenatal vs. postnatal gene sets.

Some gene sets were downloaded here https://github.com/BreenMS/Rapid-gene-set-enrichment


```{r fig.height=5, fig.width=9, warning=F, message=FALSE}

gene_pp_bf_posadded_origin <- readRDS("/project/xinhe/xsun/rare_variants/5.revision/results/mirage_res/gene_pp_bf_posadded_origin.RDS")

get_fisher_enrichment <- function(category_genes, category_name) {
  # Restrict all sets to background to be consistent
  bg <- background_genes
  highpip <- intersect(highpip_genes, bg)
  category <- intersect(category_genes, bg)
  
  # Counts for contingency table
  a <- sum(highpip %in% category)              # HighPIP & In Category
  b <- sum(highpip %notin% category)           # HighPIP & Not in Category
  c <- sum((bg %notin% highpip) & (bg %in% category))  # Not HighPIP & In Category
  d <- sum((bg %notin% highpip) & (bg %notin% category)) # Not HighPIP & Not in Category
  
  contingency_table <- matrix(c(a, b, c, d), nrow = 2,
                              dimnames = list(
                                HighPIP = c("Yes", "No"),
                                InCategory = c("Yes", "No")
                              ))
  
  fisher_result <- fisher.test(contingency_table, alternative = "greater")
  
  cat("\n", category_name, "\n")
  print(contingency_table)
  print(fisher_result)
}

```

### GTEx brain tissue specific genes

We did Fisher exact test to check if the genes with PP>0.5 are enriched in the tissue-specific gene sets. 

```{r fig.height=5, fig.width=9, warning=F, message=FALSE}

merged_df <- readRDS("/project/xinhe/xsun/rare_variants/5.revision/results/validation_gtex_brain_origin.RDS")

DT::datatable(merged_df,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;',' Brain tissue upregulated genes and overlap with MIRAGE genes'),options = list(pageLength = 10) )

merged_df <- readRDS("/project/xinhe/xsun/rare_variants/5.revision/results/validation_gtex_nonbrain_origin.RDS")


DT::datatable(merged_df,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;',' non-Brain tissue upregulated genes and overlap with MIRAGE genes'),options = list(pageLength = 10) )

```

### prenatal, postnatal, unbiased gene sets

We did Fisher exact test to check if the genes with PP>0.5 are enriched in the prenatal/postnatal/unbiased gene sets. 

```{r fig.height=5, fig.width=9, warning=F, message=FALSE}

braineffect <-  readxl::read_excel("/project/xinhe/xsun/rare_variants/5.revision/data/brain_de/BrainSpan_FetalEffect.xlsx")

prenatal <- braineffect[braineffect$FetalEffect == "Prenatal",]
unbiased <- braineffect[braineffect$FetalEffect == "Unbiased",]
postnatal <- braineffect[braineffect$FetalEffect == "Postnatal",]

print(paste0("Prenatal gene number =", nrow(prenatal)))
print(paste0("Postnatal gene number =", nrow(postnatal)))
print(paste0("Unbiased gene number =", nrow(unbiased)))

print("Genes with PP > 0.5")
highpip <- gene_pp_bf_posadded_origin[gene_pp_bf_posadded_origin$PP > 0.5,]

# Define sets
highpip_genes <- highpip$Gene
prenatal_genes <- prenatal$GeneSymbol
postnatal_genes <- postnatal$GeneSymbol
unbiased_genes <- unbiased$GeneSymbol

background_genes <- gene_pp_bf_posadded_origin$Gene


# Define `%notin%` for convenience
`%notin%` <- Negate(`%in%`)

print("Prenatal")
print(highpip$Gene[highpip$Gene %in% prenatal$GeneSymbol])
get_fisher_enrichment(prenatal_genes, "Prenatal")

print("Postnatal")
print(highpip$Gene[highpip$Gene %in% postnatal$GeneSymbol])
get_fisher_enrichment(postnatal_genes, "Postnatal")

print("Unbiased")
print(highpip$Gene[highpip$Gene %in% unbiased$GeneSymbol])
get_fisher_enrichment(unbiased_genes, "Unbiased")


```




### Single cell expression data, 

To assess whether the ASD genes are enriched in specific cell types (clusters), we used a background set of 7,867 genes expressed in the experiment. For each cell type, they built a 2×2 table comparing the number of ASD genes expressed versus not expressed, and the number of non-ASD genes expressed versus not expressed. Enrichment was evaluated using odds ratios and Fisher’s exact test.

Within each cell type cluster, a gene was considered expressed if one or more of its transcripts were detected in 25% or more cells

```{r fig.height=5, fig.width=9, warning=F, message=FALSE}

fisher_res <- readRDS("/project/xinhe/xsun/rare_variants/5.revision/results/validation_scrna_fet_origin.RDS")

DT::datatable(fisher_res,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Fisher results for the ASD genes'),options = list(pageLength = 10) )

cluster_member <- readRDS("/project/xinhe/xsun/rare_variants/5.revision/results/validation_scrna_expr_binary_origin.RDS")

DT::datatable(cluster_member,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Cluster membership for the ASD genes'),options = list(pageLength = 10) )

```



