---
title: "Annotating rare variants"
author: "XSun"
date: "2025-06-19"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

We annotate the rare variants we collected using reviewer recommended scores 

```{r}
library(data.table)
library(ggplot2)

mirage_input <- fread("/project/xinhe/xsun/rare_variants/5.revision/data/mirage_input_20231013.txt",header = T)

#DT::datatable(mirage_input,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','MIRAGE input -- initial submission'),options = list(pageLength = 10) )

group_labels <- c("highPLI", "medPLI", "lowPLI", "highMPC", "medMPC", "lowMPC")
mirage_input[, group := group_labels[group.index]]

print('MIRAGE input -- initial submission')
print(head(mirage_input))

table(mirage_input$group)[group_labels]
```


# Alpha-missense score

Alpha-missense score download: https://console.cloud.google.com/storage/browser/dm_alphamissense;tab=objects?pli=1&inv=1&invt=Ab0dvQ&prefix=&forceOnObjectsSortingFiltering=false

Paper: https://www.science.org/doi/10.1126/science.adg7492

The authors compared with alphamissense with other missense annotations: https://www.ebi.ac.uk/training/online/courses/alphafold/classifying-the-effects-of-missense-variants-using-alphamissense/introducing-missense-variants-and-alphamissense/

```
zcat /project/xinhe/xsun/rare_variants/5.revision/data/alphamis/AlphaMissense_hg38.tsv.gz |head
# Copyright 2023 DeepMind Technologies Limited
#
# Licensed under CC BY-NC-SA 4.0 license
CHROM  POS     REF     ALT     genome  uniprot_id      transcript_id   protein_variant am_pathogenicity        am_class
chr1    69094   G       T       hg38    Q8NH21  ENST00000335137.4       V2L     0.2937  likely_benign
chr1    69094   G       C       hg38    Q8NH21  ENST00000335137.4       V2L     0.2937  likely_benign
chr1    69094   G       A       hg38    Q8NH21  ENST00000335137.4       V2M     0.3296  likely_benign
chr1    69095   T       C       hg38    Q8NH21  ENST00000335137.4       V2A     0.2609  likely_benign
chr1    69095   T       A       hg38    Q8NH21  ENST00000335137.4       V2E     0.2922  likely_benign
```

```
| Category          | Min pathogenicity | Max pathogenicity | Total Variants |
|-------------------|-----------|-----------|----------------|
| likely_pathogenic | 0.5640    | 1.0000    | 22,770,557     |
| likely_benign     | 0.0001    | 0.3400    | 40,917,351     |
| ambiguous         | 0.3400    | 0.5640    | 8,009,648      |
```


```{r}
mirage_amsocre <- readRDS("/project/xinhe/xsun/rare_variants/5.revision/data/mirage_input_mislof_added.RDS")
mirage_amsocre <- as.data.table(mirage_amsocre)
mirage_amsocre <- mirage_amsocre[,-c("key")]
mirage_amsocre[, old_group := group_labels[group.index]]

# oldlofmisnow <- mirage_amsocre[!is.na(am_class) & group.index %in% 1:3]
# 
# DT::datatable(oldlofmisnow,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Variants annotated as LoF before but missense now'),options = list(pageLength = 5) )

oldmisnanow <- mirage_amsocre[is.na(am_class) & group.index %in% 4:6]

DT::datatable(oldmisnanow,caption = htmltools::tags$caption( style = 'caption-side: left; text-align: left; color:black;  font-size:150% ;','Variants annotated as missense before but have no am_pathogenicity now'),options = list(pageLength = 5) )

mirage_amsocre[!is.na(am_class), old_group := am_class]
table(mirage_amsocre$old_group)[c("highPLI","medPLI","lowPLI","likely_pathogenic","ambiguous","likely_benign")]

```


# REVEL

REVEL score download: https://sites.google.com/site/revelgenomics/downloads?authuser=0

Paper: https://pubmed.ncbi.nlm.nih.gov/27666373/

```
head /project/xinhe/xsun/rare_variants/5.revision/data/revel/revel_with_transcript_ids
chr,hg19_pos,grch38_pos,ref,alt,aaref,aaalt,REVEL,Ensembl_transcriptid
1,35142,35142,G,A,T,M,0.027,ENST00000417324
1,35142,35142,G,C,T,R,0.035,ENST00000417324
1,35142,35142,G,T,T,K,0.043,ENST00000417324
1,35143,35143,T,A,T,S,0.018,ENST00000417324
1,35143,35143,T,C,T,A,0.034,ENST00000417324
1,35143,35143,T,G,T,P,0.039,ENST00000417324
1,35144,35144,A,C,C,W,0.012,ENST00000417324
1,35145,35145,C,A,C,F,0.023,ENST00000417324
1,35145,35145,C,G,C,S,0.029,ENST00000417324

```

82,100,678 variants in total


# LOEUF

LOEUF paper https://www.nature.com/articles/s41586-020-2308-7

Data download https://gnomad.broadinstitute.org/downloads#v4-constraint under constraint

**Differences between LOEUF and PLI**

https://gnomad.broadinstitute.org/help/constraint#loeuf-vs-pli


```{r fig.height=3, fig.width=5}

ggplot(mirage_amsocre[!is.na(am_pathogenicity)], aes(x = am_pathogenicity)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "AlphaMissense Pathogenicity Scores for all MIRAGE variants",
       x = "AlphaMissense Pathogenicity",
       y = "Count")

```



